---
title: "Genetically Proxied Lipid Therapy And Sepsis"
author: "Nicholas Sunderland"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genetically Proxied Lipid Therapy And Sepsis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
  wrap: 80
---

```{r init, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>", 
  eval     = FALSE,
  cache    = FALSE
)
```

## Installation
You can install the latest version of MRLipidInfection in R using:
```{r installation, eval = FALSE}
remotes::install_github("nicksunderland/MRLipidInfection")
```

## Setup
First we need to load the MRLipidInfection package and setup package requirements by calling
`setupMRLipidInfection()`. At the first installation this will take some time as this will download
the [1000 genomes LD reference panel](http://fileserve.mrcieu.ac.uk/ld/1kg.v3.tgz) into the package
file structure and ensure that [PLINK](https://www.cog-genomics.org/plink/) is installed locally.

```{r setup}
library(MRLipidInfection)

setupMRLipidInfection()
```

<br>

# Exposure data
## Lipid profile data
Next we must obtain the data for the exposure. We are interested whether exposure to lipid lowering
medications influences risk of developing sepsis. This data is not readily available on a population
level and has many confounding factors. We will therefore estimate the effect of genetically proxied
lipid medication therapy by looking for Single Nucleotide Polymorphisms (SNPs) in close proximity to
known lipid metabolism genes (e.g. HMG-CoA reductase) that are significantly associated with
circulating lipid levels. We will use the the latest Global Lipids Genetics Consortium
[Ancestry-specific GWAS summary statistics](http://csg.sph.umich.edu/willer/public/glgc-lipids2021/)
for HDL-C, LDL-C, nonHDL-C, TC and TG. 

First we will download the data for LDL cholesterol - **Warning** GWAS data file is large and will
take some time to download.

```{r download_data}
# The URL for the GWAS summary statistics README document
url_rm <- "http://csg.sph.umich.edu/willer/public/glgc-lipids2021/results/ancestry_specific/README.txt"

# Download README to the package downloads directory (subdirectory "glgc").
get_web_file(url_rm, directory="glgc")

# The URLs for the GWAS summary statistics data
lipids_gwas_urls <- list(
  ldl    = "http://csg.sph.umich.edu/willer/public/glgc-lipids2021/results/ancestry_specific/without_UKB_LDL_INV_EUR_HRC_1KGP3_others_ALL.meta.singlevar.results.gz.tbi",
  hdl    = "http://csg.sph.umich.edu/willer/public/glgc-lipids2021/results/ancestry_specific/without_UKB_HDL_INV_EUR_HRC_1KGP3_others_ALL.meta.singlevar.results.gz.tbi",
  tc     = "http://csg.sph.umich.edu/willer/public/glgc-lipids2021/results/ancestry_specific/without_UKB_TC_INV_EUR_HRC_1KGP3_others_ALL.meta.singlevar.results.gz.tbi",
  nonhdl = "http://csg.sph.umich.edu/willer/public/glgc-lipids2021/results/ancestry_specific/without_UKB_nonHDL_INV_EUR_HRC_1KGP3_others_ALL.meta.singlevar.results.gz.tbi",
  tg     = "http://csg.sph.umich.edu/willer/public/glgc-lipids2021/results/ancestry_specific/without_UKB_logTG_INV_EUR_HRC_1KGP3_others_ALL.meta.singlevar.results.gz.tbi"
  )

# Download GLGC GWAS data to package downloads directory; return the file paths
lipids_gwas_fps <- purrr::map(lipids_gwas_urls, get_web_file, directory="glgc")
lipids_gwas_fps
```

There are more than 47 million SNPs in the dataset. We now filter the GWAS data SNPs associated with
circulating LDL cholesterol levels at a statisticsal significance of `p<5e-8`. Reading this file can
take a long time and so we can write out the result to a .csv file.

```{r significant_snps}
# set significance level for filtering SNPs 
p <- 5e-8

# Read the .gz data files and filter by significance level
sig_snps <- purrr::map2(.x = lipids_gwas_fps,                   
                        .y = names(lipids_gwas_fps),  
                        .f = ~extract_significant_snps(.x, .y, pval=p, pval_col="pvalue", snp_col="rsID"))

# View one of the tables:
utils::head(sig_snps$ldl)

# View the file path to the data in .csv format
attr(sig_snps$ldl, "filepath")

```

If we inspect the column names we see that they need some adjustment to ensure they are compatible
with the TwoSampleMR package naming conventions.
```{r standardise}
# The current column names
colnames(sig_snps$ldl)

# Standardise exposure data column names
exposure_data <- purrr::map(.x = sig_snps,
                            .f = ~ TwoSampleMR::format_data(
                                    dat               = .x,
                                    type              = "exposure",
                                    phenotype_col     = "phenotype",
                                    snp_col           = "rsID",
                                    beta_col          = "EFFECT_SIZE",
                                    se_col            = "SE",
                                    eaf_col           = "POOLED_ALT_AF",
                                    effect_allele_col = "ALT",
                                    other_allele_col  = "REF",
                                    pval_col          = "pvalue",
                                    samplesize_col    = "N",
                                    chr_col           = "CHROM",
                                    pos_col           = "POS_b37"))
```

## Cis-acting HMG-CoA Reductase SNPs
We will create a second set of exposure SNPs that will be subset of the SNPs that associate with LDL
cholesterol levels. This subset of SNPs will be within a certain proximity (number of bases) to the
HMG-CoA Reductase gene. Data regarding the position of the HMG-CoA-R gene is freely available
[online](https://www.genecards.org/cgi-bin/carddisp.pl?gene=HMGCR). The Lipid data from the GLGC
uses the Genome Reference Consortium Human Reference 37 (GRCh37) and so we must use this when
referring to the position of the HMG-CoA-R gene.

```{r cis_hmg}
# The HMG-CoA-R, NPC1L1, and PCSK9 genes; set the 'cis' base tolerance here.
genes_list = list(
  Gene(name="HMGCR",  chromosome=5, start=74632993, end=74657941, cis_tol=5e5),
  Gene(name="NPC1L1", chromosome=7, start=44552134, end=44580929, cis_tol=5e5),
  Gene(name="PCSK9",  chromosome=1, start=55505221, end=55530525, cis_tol=5e5)
)

# Create a new dataset of SNPs close to the genes
cis_exp_dat <- create_cis_gene_exposure(exposure_data, genes_list)
```

## Clumping SNPs
Many of the SNPs will be highly correlated, i.e. be genetically linked and provide no additional
information over and above the whole set of correlated SNPs. To reduce the number of SNPs to only
those most significantly associated with lipid levels we use Linkage Disequilibrium clumping.
[Clumping](https://kp4cd.org/sites/default/files/documents/LD_clumping.pdf) reports the most
significant genetic associations in a region in terms of a smaller number of “clumps” of genetically
linked SNPs.

SNPs that are correlated with an R^2^ greater than some value `clump_r2` will be clumped - i.e. the
SNP with the lower p-value will be removed. The process is iterative and has several parameters:
`clump_kb`, `clump_r2`, `clump_p`. We will use the defaults, except for `clump_r2` with the
cis-HMG-CoA-R SNPs, which we will set to 0.1 instead of the default of 0.001.

We could do this on the Bristol IEU severs using the function `TwoSampleMR::clump_data`. 
```{r clumping_1}
# Clump the SNPs 
clumped_exp_dat <- clump_snps(cis_exp_dat, clump_r2=0.2)

```

<br>

# Outcome data
## Sepsis
The outcome we are interested in is sepsis. There are several GWAS relating whether a patient
develops sepsis to their underlying SNPs. Several of these studies are available on the [IEU GWAS
database](https://gwas.mrcieu.ac.uk) which can be queried using either the `TwoSampleMR` or
`ieugwasr` R packages.
```{r outcome}
# List of available GWAS summary statistics
available_gwas <- ieugwasr::gwasinfo() |>
  
  # Filter for those that mention 'sepsis' in the 'trait' description column
  dplyr::filter(grepl("sepsis|pneumonia", .data$trait, ignore.case=TRUE))

# We will pick one
outcome_gwas_id <- "ieu-b-4980"

# View
available_gwas[available_gwas$id == outcome_gwas_id, ]
```
## Extracting outcome data
Having identified the GWAS summary statistics we next need to extract the data for our SNPs of
interest (those that are highly predictive of the exposure, in this case LDL cholesterol).
```{r outcome_extracting}
# The cisHMG-CoA SNPs won't necessarily all be within the group of SNPs for the whole genome.
##all(clumped_hmg_exp$SNP %in% clumped_exp$SNP)

# Extract data for cis-HMG-CoA SNPs
outcome_data <- purrr::map(.x = clumped_exp_dat,
                           .f = ~ TwoSampleMR::extract_outcome_data(
                              snps            = .x$SNP,
                              outcomes        = outcome_gwas_id,
                              proxies         = TRUE,
                              rsq             = 0.8,
                              align_alleles   = 1,
                              palindromes     = 1,
                              maf_threshold   = 0.3,
                              access_token    = ieugwasr::check_access_token(),
                              splitsize       = 10000,
                              proxy_splitsize = 500))
```
## Obtaining outcome data _not_ on the IEU GWAS database
How to extract outcome data from other sources and find proxy SNPs
```{r outcome_other_source}
```

<br>

# Harmonising
We have the effects of the (significant) SNPs on LDL levels and the effect of these same SNPs on the
outcome sepsis. However, it is important to harmonise the effects. This means that the effect of a
SNP on the exposure and the effect of that SNP on the outcome must each correspond to the same
allele (i.e. the A&Ts and C&Gs cannot be the wrong way round). 
*Note* The IEU GWAS database contains data that is already harmonised. For more information on
[harmonising](https://mrcieu.github.io/TwoSampleMR/articles/harmonise.html).
```{r harmonise}
# Filter out the Gene:Lipid SNP associations with no significant SNPs
zero_snps <- purrr::map_lgl(clumped_exp_dat, ~ nrow(.x)==0)

# Harmonise the data for the LDL SNPs across the whole genome
harm_dat <- purrr::map2(.x = clumped_exp_dat[!zero_snps],
                        .y = outcome_data[!zero_snps],
                        .f = ~ TwoSampleMR::harmonise_data(.x, .y))
```   

<br>

# Two Sample Mendelian Randomisation
The whole point is to assess the weight influence of the SNPs on the outcome
```{r mr}
# The MR result using SNPs from across the genome
res_dat <- purrr::map(.x = harm_dat, 
                      .f = ~ TwoSampleMR::mr(.x))
```

## Plotting the result
```{r plot}
plots <- purrr::map2(.x = res_dat, 
                     .y = harm_dat,
                     .f = ~ TwoSampleMR::mr_scatter_plot(.x, .y))
plots
```

## To be continued
